Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-11-27 threshold/draft.Rnw
minor tweaks to algo, working on efficiency now...

The number of correlation coefficients grows quadratically with the number of
vectors. Simply computing all pairs of correlation coefficients and then
filtering out coefficients below a given threshold may not be computationally
feasable for high-dimensional data in modern genomics and other applications.
Considerable attention has been devoted to \emph{pruning
methods} that cheaply prune away all but the most highly-correlated pairs of
vectors.  For example the TAPER and related algorithms of Xiong et al. develop
pruning rules based on the frequency of mutually co-ocurring observations (rows)
among vector pairs; see \cite{prune1, prune2} and the references therein.
Related methods approximate a set containing the most highly-correlated
pairs using hashing algorithms\cite{prune3}. A number of other methods arrive
at approximate sets of the most highly-correlated pairs using
dimensionality-reduction techniques like the discrete Fourier
transform\cite{timeseries}. The method of Wu, et al.\cite{svd-similarity},
conceptually very similar to our approach, uses the singular value
decomposition (SVD) to find all pairs of close vectors with respect to a
distance metric (instead of correlation).  Our method finds all of the most
highly-correlated vector pairs with respect to a given threshold by pruning
along an intuitive path of decreasing variance defined by the data and computed
by the SVD.

@@

Algorithm \ref{prune} guarantees that no pruned vector pair exceeds the given
correlation threshold--the pruning does not produce false negatives.
The algorithm typically prunes the vast majority of pairs of vectors below the
correlation threshold with truncated singular value decompositions of only
relatively low rank $p$.  The choice of $p$ represents a balance between work in
computing a truncated SVD and work required by the $\ell$ matrix vector
products of order $n\times p$ in step 4.

>>>>

The number of correlation coefficients grows quadratically with the number of
vectors. Simply computing all pairs of correlation coefficients and then
filtering out coefficients below a given threshold may not be computationally
feasable for high-dimensional data in modern genomics and other applications.
Considerable attention has been devoted to \emph{pruning
methods} that cheaply prune away all but the most highly-correlated pairs of
vectors.  For example the TAPER and related algorithms of Xiong et al. develop
pruning rules based on the frequency of mutually co-ocurring observations (rows)
among vector pairs; see \cite{prune1, prune2} and the references therein.
Related methods approximate a set containing the most highly-correlated
pairs using hashing algorithms\cite{prune3}. A number of other methods arrive
at approximate sets of the most highly-correlated pairs using
dimensionality-reduction techniques like the discrete Fourier
transform\cite{timeseries}. The method of Wu, et al.\cite{svd-similarity},
conceptually very similar to our approach, uses the singular value
decomposition (SVD) to find all pairs of close vectors with respect to a
distance metric (instead of correlation).  Our method finds all of the most
highly-correlated vector pairs with respect to a given threshold by pruning
along an intuitive path of decreasing variance defined by the data and computed
by a truncated SVD.

@@

Algorithm \ref{prune} guarantees that no pruned vector pair exceeds the given
correlation threshold--the pruning does not produce false negatives.
The algorithm typically prunes the vast majority of pairs of vectors below the
correlation threshold with truncated singular value decompositions of only
relatively low rank $p$.  The choice of $p$ represents a balance between work in
computing a truncated SVD plus work required by step 4 and filtering the
thresholded values in step 7.
